Multi-target detection with application to cryo-electron microscopy

### 2019-03-01 inverse_problems/inverse_problems.tex
Section 4 (first part)

Owing to the separation condition, when correlating $y$ with shifted versions of itself for shifts in $0, \ldots, L-1$, any given occurrence of $x$ in $y$ is only ever correlated with itself, and never with another occurrence. As a result, the autocorrelations of $y$ depend on the corresponding autocorrelations of $x$, the noise level $\sigma$ and the density $\gamma$ (which is a weak dependence on the support signal $s$). Specifically, we show the following identities in Section~\ref{sec:autocorrelation_computation}:
\begin{align}
	a_y^1 & = \gamma a_{x}^1, \label{eq:mean_micrograph} \\
	a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell], \label{eq:ac2_micrograph} \\
	a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2]  + \sigma^2\gamma a_{x}^1  \big(\delta[\ell_1]+\delta[\ell_2]
	+\delta[\ell_1-\ell_2]\big), \label{eq:ac3_micrograph}
\end{align}
where $\delta[0] = 1$ and $\delta[\ell \neq 0] = 0$, and indices $\ell, \ell_1, \ell_2$ are in the range $0 \leq \ell \leq L-1$. Terms proportional to $\sigma^2$ are due to the noise. If $\sigma$ is known, they can be handled easily. If $\sigma$ is unknown, one can either estimate it form the data, or one can ignore the few entries of the autocorrelations that are affected by $\sigma$---one in $a_y^2$ and $3L-2$ in $a_y^3$, a relatively small number in both cases.

@@

A signal is determined uniquely by its second- and third-order autocorrelations. Indeed, assuming $z[0]$ and $z[L-1]$ are nonzero (otherwise, redefine the length of the signal), we can recover $z$ explicitly using this identity for $k = 0, \ldots, L-1$:
%
\begin{equation}
%
z[k]  = \frac{z[0]z[k]z[L-1]}{z[0]z[L-1]} = \frac{a_z^3[k,L-1]}{a_z^2[L-1]}.
\label{eq-uniqueness}
%
\end{equation}
This proves the following useful fact:
\begin{proposition} \label{prop:uniqueness}
	%
	A signal $z\in\RL$ is determined uniquely from  $a_z^2$ and $a_z^3$. 
\end{proposition}

Note that~\eqref{eq-uniqueness} is not numerically stable: if $z[0]$ or $z[L-1]$ are close to 0, recovery of $z$ is sensitive to errors in the autocorrelations. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedure, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.

@@

The observed moments $a_y^1,a_y^2$ and $a_y^3$ of $y$ do not immediately yield the moments of the signal $x$, as seen by~\eqref{eq:mean_micrograph},~\eqref{eq:ac2_micrograph} and~\eqref{eq:ac3_micrograph}; rather, the two are related by the noise level $\sigma$ and the ratio $\gamma$. We will show, however, that $x$ is still determined by the observed moments of $y$.

First, we observe that if the noise level $\sigma$ is known, generally, one can estimate $\gamma$ from the first two moments of the micrograph. The proof is provided in Appendix~\ref{sec:proof_prop_gamma}.

@@

\begin{align}
\gamma = L \frac{(\gamma a_x^1) (\gamma a_x^2[1])}
{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
	+ \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]}.
\end{align}
%With $\gamma$ in hand, we can estimate $\sigma^2$:
%
\begin{align}
\sigma^2 = a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
- (2 L - 1) (a_y^1)^2.
\end{align}

>>>>

Owing to the separation condition, when correlating $y$ with shifted versions of itself for shifts in $0, \ldots, L-1$, any given occurrence of $x$ in $y$ is only ever correlated with itself, and never with another occurrence. As a result, the autocorrelations of $y$ depend on the corresponding autocorrelations of $x$, the noise level $\sigma$ and the density $\gamma$ (which is a weak dependence on the support signal $s$). Specifically, we show the following identities in Section~\ref{sec:autocorrelation_computation}:
\begin{align} \label{eq:ac_well_separated}
	a_y^1 & = \gamma a_{x}^1, \nonumber \\
	a_y^2[\ell] & = \gamma a_{x}^2[\ell] + \sigma^2\delta[\ell], \\
	a_y^3[\ell_1,\ell_2] & = \gamma a_{x}^3[\ell_1,\ell_2]  + \sigma^2\gamma a_{x}^1  \big(\delta[\ell_1]+\delta[\ell_2]
	+\delta[\ell_1-\ell_2]\big), 
\end{align}
where $\delta[0] = 1$ and $\delta[\ell \neq 0] = 0$, and indices $\ell, \ell_1, \ell_2$ are in the range $0 \leq \ell \leq L-1$. Terms proportional to $\sigma^2$ are due to the noise. If $\sigma$ is known, they can be handled easily. If $\sigma$ is unknown, one can either estimate it form the data, or one can ignore the few entries of the autocorrelations that are affected by $\sigma$---one in $a_y^2$ and $3L-2$ in $a_y^3$, a relatively small number in both cases.

@@

A signal $z\in\RL$ is determined uniquely by its second- and third-order autocorrelations. Indeed, assuming $z[0]$ and $z[L-1]$ are nonzero (otherwise, redefine the length of the signal), we can recover $z$ explicitly using this identity for $k = 0, \ldots, L-1$:
%
\begin{equation}
%
z[k]  = \frac{z[0]z[k]z[L-1]}{z[0]z[L-1]} = \frac{a_z^3[k,L-1]}{a_z^2[L-1]}.
\label{eq-uniqueness}
%
\end{equation}
This proves the following useful fact:
\begin{proposition} \label{prop:uniqueness}
	%
	A signal $z\in\RL$ is determined uniquely from  $a_z^2$ and $a_z^3$. 
\end{proposition}

Note that the procedure described in~\eqref{eq-uniqueness} is not numerically stable: if $z[0]$ or $z[L-1]$ are close to 0, recovery of $z$ is sensitive to errors in the autocorrelations. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedure, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.

In what follows, we show that in the homogeneous case $K=1$, the signal $x$, the parameter $\gamma$ and $\sigma^2$ are identified uniquely from the first three observed autocorrelations under both the well-separated and Poisson models. Then, we bound the maximal number of signals that can be recovered from  those autocorrelations in the heterogeneous case, and discuss extensions to multi-dimensional signals.

@@

The observed moments $a_y^1,a_y^2$ and $a_y^3$ under the well-separated model~\eqref{eq:ac_well_separated} do not immediately yield the autocorrelations  of the signal $x$; rather, the two are related by the noise level $\sigma$ and the ratio $\gamma$. We will show, however, that $x$ is still determined by the observed moments of $y$.

First, we observe that if the noise level $\sigma$ is known, generally, one can estimate $\gamma$ from the first two moments of the micrograph. The proof is provided in Section~\ref{sec:proof_prop_gamma}.

@@

\begin{align}
\gamma &= L \frac{(\gamma a_x^1) (\gamma a_x^2[1])}
{\sum_{\ell=0}^{L-1} \gamma a_x^3[1,\ell] 
	+ \sum_{\ell=1}^{L-2} \gamma a_x^3[\ell,\ell+1]}.
\end{align}
%With $\gamma$ in hand, we can estimate $\sigma^2$:
%
\begin{align}
\sigma^2 &= a_y^2[0] + 2\sum_{\ell = 1}^{L-1}a_y^2[\ell]-\frac{L (a^1_y)^2}{\gamma}
- (2 L - 1) (a_y^1)^2.
\end{align}
